---
title: "CellType: identifying the celltype by marker genes"
date: "Created: 2021-02-01; modified: `r Sys.Date()`"
output: 
  html_notebook:
    codes: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(Seurat)
library(tibble)
library(ggplot2)
library(cowplot)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 4000 * 1024^2)

colors <- c("#A6CEE3", "#1F78B4", "#08306B", "#B2DF8A", "#006D2C", "#8E0152",
         "#DE77AE", "#CAB2D6", "#6A3D9A", "#FB9A99", "#E31A1C", "#B15928",
         "#619CFF","#FF67A4","#00BCD8", "#EE2B2B", "#2D6BB4")
```


### load data 
```{r}
data.cluster <- readRDS("../../Result/sc_profile/hsp_normal/data.normal.cluster.RDS")
```


### Segregation of clusters by group
```{r}
n_cells <- FetchData(data.cluster, 
                     vars = c("ident", "Batch")) %>%
        dplyr::count(ident, Batch) %>%
        tidyr::spread(ident, n)
head(n_cells)
# DT::datatable(n_cells)
```


### visualization
```{r, fig.width=12, fig.height=6}
cowplot::plot_grid( 
        DimPlot(data.cluster,
        reduction = "umap",
        label = TRUE,
        label.size = 6), 
        DimPlot(data.cluster,
        reduction = "umap",
        label = TRUE,
        label.size = 6, 
        split.by = "Batch")+NoLegend(),
        ncol = 2)
```

* Segregation of clusters by various sources of uninteresting variation

```{r}
metrics <-  c("nFeature_SCT", "percent.mt", "S.Score", "G2M.Score")
FeaturePlot(data.cluster, 
            reduction = "umap", 
            features = metrics,
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

### UMAP for clusters
```{r}
colors2 <- c("#A6CEE3", "#1F78B4", "#08306B", "#B2DF8A", "#006D2C", 
            "#8E0152","#DE77AE", "#CAB2D6", "#6A3D9A", "#FB9A99", 
            "#E31A1C", "#B15928", "#619CFF","#FF67A4","#00BCD8", 
            "#EE2B2B", "#2D6BB4","#AFD300", "#B089D8", "#F96E6F",
            "#AAD3ED", "#639BCE")

# data.cluster.v2 <- data.cluster
# Idents(object = data.cluster.v2) <- "integrated_snn_res.0.4"

DimPlot(data.cluster,
        reduction = "umap",
        label = TRUE,
        label.size = 8,
        cols = colors2)
```


### Exploration of the PCs driving the different clusters
```{r}
# Defining the information in the seurat object of interest
columns <- c(paste0("PC_", 1:16),
            "ident",
            "UMAP_1", "UMAP_2")

# Extracting this data from the seurat object
pc_data <- FetchData(data.cluster, 
                     vars = columns)

# Adding cluster label to center of cluster on UMAP
tsne_label <- FetchData(data.cluster, 
                        vars = c("ident", "UMAP_1", "UMAP_2"))  %>%
  group_by(ident) %>%
  summarise(x=mean(UMAP_1), y=mean(UMAP_2))
  
# Plotting a UMAP plot for each of the PCs
pc_plot <- purrr::map(paste0("PC_", 1:16), function(pc){
        ggplot(pc_data, 
               aes(UMAP_1, UMAP_2)) +
                geom_point(aes_string(color=pc), 
                           alpha = 0.7) +
                scale_color_gradient(guide = FALSE, 
                                     low = "grey90", 
                                     high = "blue")  +
                geom_text(data=tsne_label, 
                          aes(label=ident, x, y)) +
                ggtitle(pc)
}) %>% plot_grid(plotlist = .)
```

* Examine PCA results 
```{r}
print(data.cluster[["pca"]], dims = 1:5, nfeatures = 5)
```



### Identification of all markers for each cluster
```{r}
annotations <- read.csv("../../Result/sc_profile/Celltype/annotation.csv")
if(0){
  
  # markers <- FindAllMarkers(object = data.cluster,
  #                           only.pos = TRUE,
  #                           logfc.threshold = 0.25)
  
  markers <- FindAllMarkers(object = data.cluster,
                            test.use = "wilcox",
                            min.pct = 0.2,
                            min.diff.pct = 0.1,
                            return.thresh = 0.05,
                            only.pos = TRUE,
                            logfc.threshold = 0.25) %>%
    filter(p_val_adj < 1e-06)
  write.csv(markers, "../../Result/sc_profile/hsp_normal/data.normal.markergenes.csv") 
}

markers <- read.csv("../../Result/sc_profile/hsp_normal/data.normal.markergenes.csv", row.names = 1)
```


### evaluating marker genes : identify the top gene marker per cluster
```{r, fig.width=12, fig.height=10}
anno_all_markers <- markers %>%
        left_join(y = unique(annotations[, c("gene_name", "description")]),
                  by = c("gene" = "gene_name")) %>%
        mutate(Specific=pct.1 - pct.2) #%>%
  #arrange(dplyr::desc(Specific), dplyr::desc(avg_logFC))

# Extract top markers per cluster
topNumber <- anno_all_markers %>% 
    group_by(cluster) %>% 
    top_n(n = 20, wt = avg_logFC) %>%
    ungroup()
mapal <- colorRampPalette(c("#EEEED1", "#AEDCF9", "#10106A"))(20)
top_heatmap <- DoHeatmap(object = data.cluster, 
                         features = topNumber$gene, 
                         label = TRUE)+
  scale_fill_gradientn(colours = mapal)
top_heatmap
```


### pearson correlation

choosing the average expression per cluster
```{r}
library(pheatmap)
exp_ave <- AverageExpression(data.cluster)
CorOb.cor.exp <- as.data.frame(cor(exp_ave$SCT, method = "spearman"))
pheatmap(CorOb.cor.exp)
```


### The marker genes of Heart cell types 

#### 1. cardiomyocyte

  * TTN, MYH7, MYH6, TNNT2,  
  
  * MLCK2, MYLK1, AAT7, MYLK 
  
  * BNP, NPPA, NPPB,  
  
  * ACTC1, TPM, RYR2, ATP2A2, TNNC1, ACTA1
```{r, fig.width=8, fig.height=6}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            # features = c("TTN", "MYH7", "MYH6", "TNNT2", "MLCK2", "MYLK1", "AAT7", "MYLK", 
            #              "BNP", "NPPA", "NPPB", "ACTC1", "TPM", "RYR2", "ATP2A2", "TNNC1", "ACTA1"),
            features = c("MYH7", "MYH6", "MYLK", "NPPA", "NPPB", "ACTA1"),            
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

**Results: cluster 1, 5, 7, 8 (maybe include cluster 6, 12)**


#### 2. fibroblasts: 

  * LUM, DCN, C7, SERPINF1, FBLN1
  
  * COL1A2, COL3A1, VIM, FSTL1, GSN, FBLN2, SPARC, MMP2
```{r, fig.width=10, fig.height=8}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            features = c("LUM", "DCN", "C7", "SERPINF1", "FBLN1",  
                         "COL1A2", "COL3A1", "VIM", "FSTL1", "GSN", "FBLN2", "SPARC", "MMP2"),
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

**Results: cluster 0, 3, 10, 13 (maybe include cluster 6, 12)**


#### 3. endothelial cell: 

  * AQP1, BST2, CXCL2, VWF, IFI27, PECAM1
  
  * FABP4, TIE1, EGFL7, FLT1, EPAS, EMCN, EDNRB
```{r, fig.width=10, fig.height=6}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            # features = c("AQP1", "CXCL2","VWF", "PECAM1", "IFI27",
            #              "FABP4", "TIE1", "EGFL7", "FLT1", "EPAS", "EMCN", "EDNRB"),
            features = c("AQP1", "CXCL2","VWF", "PECAM1", "IFI27",
                         "FABP4", "EGFL7", "FLT1", "EMCN", "EDNRB"),
            pt.size = 0.2, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

**Result: cluster 0, 10, 13**

#### 4. smooth muscle cell: 

  * MYH11, CALD1, ACTA2, IGFBP7, RGS5 
```{r}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            features = c("MYH11", "CALD1", "ACTA2", "RGS5"),
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

**Result: cluster 6, 12**


#### 5. Macrophages: 

  * AIF1, MS4A6A, C3AR1, CXCL8
  
  * CD74, CD163, LYZ1, EMR1, CSFR1, LGALS3, CD68, ITGAM
```{r, fig.width=8, fig.height=6}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            # features = c("AIF1", "MS4A6A", "C3AR1", "CXCL8", 
            #              "CD74", "CD163", "LYZ1", "EMR1", "CSFR1", "LGALS3", "CD68", "ITGAM"),
            
            features = c("AIF1", "MS4A6A", "CXCL8", "CD74", "CD163", "CD68"),            
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```

**Result: cluster 14**

### gene Markers: manual annotation

[panglaodb](https://panglaodb.se/index.html)

|   Cluster   |      Celltype           |
| :---------: | :----------------------:|
|     0       |     cardiomyocyte       |
|     1       |     endothelial cell    |
|     2       |       unknow cell       |
|     3       |    smooth muscle cell   |
|     4       |       fibroblasts       |
|     5       |       unknow cell       |
|     6       |       cardiomyocyte     |
|     7       |      cardiomyocyte      |
|     8       |        unknow cell      |
|     9       |   smooth muscle cell    |
|     10      |      macrophages        |
|     11      |     endothelial cell    |
|     12      |   smooth muscle cell    |
|     13      |      macrophages        |
|     14      |     endothelial cell    |


### Rename all identities
```{r}
data.cluster_rename <- RenameIdents(object = data.cluster, 
                               "0" = "cardiomyocyte",
                               "1" = "endothelial cells",
                               "2" = "unknow cells",
                               "3" = "smooth muscle cells",
                               "4" = "fibroblast",
                               "5" = "unknow cells",
                               "6" = "cardiomyocyte",
                               "7" = "cardiomyocyte",
                               "8" = "unknow cells",
                               "9" = "smooth muscle cells",
                               "10" = "macrophage",
                               "11" = "endothelial cells")

DimPlot(object = data.cluster_rename, 
        reduction = "umap", 
        label = TRUE,
        label.size = 5,
        repel = TRUE,
        cols = colors)
```

* Remove the stressed or dying cells
```{r}
data.cluster_rename <- subset(data.cluster_rename,
                          idents = "unknow cells", 
                          invert = TRUE)

# Re-visualize the clusters
DimPlot(object = data.cluster_rename, 
        reduction = "umap", 
        label = TRUE,
        label.size = 5,
	      repel = TRUE,
        cols = colors)
```

```{r}
n_cells2 <- FetchData(data.cluster_rename, 
                     vars = c("ident", "Individual")) %>%
        dplyr::count(ident, Individual) %>%
        tidyr::spread(ident, n)
DT::datatable(n_cells2)

n_cells2 <- FetchData(data.cluster_rename, 
                     vars = c("ident", "Batch")) %>%
        dplyr::count(ident, Batch) %>%
        tidyr::spread(ident, n)
DT::datatable(n_cells2)
```


### save data
```{r}
saveRDS(data.cluster, "../../Result/sc_profile/RDS/data.normal.celltype.rds", compress = TRUE)
saveRDS(data.cluster_rename, "../../Result/sc_profile/RDS/data.normal.celltype_rename.rds", compress = TRUE)
```


### 20210305

Investigating the different stages of Macrophages, including M1 and M2 Macrophages via marker genes

* inflammation-inducing M1 macrophages: FCGR1A, IDO1, SOCS1, CXCL10
```{r}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            features = c("FCGR1A", "IDO1", "SOCS1", "CXCL10"),
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```


* inflammation-inhibiting M2 macrophages: MRC1, TGM2, FCER2, CCL22
```{r}
FeaturePlot(data.cluster, 
            reduction = "umap", 
            features = c("MRC1", "TGM2", "FCER2", "CCL22"),
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```


### version
```{r}
sessionInfo()
```


1. [不同单细胞R包的数据格式相互转换](https://www.jianshu.com/p/50e289a694d2)

