---
title: "Enriched pathway analysis: GO term of Biological Process"
date: "Created: 2021-03-23 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

## Introduction

Performing  Over Representation Analysis of the up- or down-regulated DEGs  via **clusterProfiler** and **enrichplot** package based on the limma package.

* *clusterProfiler* calculates the enrichment score

* *enrichplot* displays the enrichment result

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(ggplot2)
library(data.table)
library(clusterProfiler)
library(enrichplot)
library(patchwork)

rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

mus.grp <- c("TAC_NC", "TAC", "HF_NC", "HF")
mus.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F")

hsp.grp <- c("NC", "DCM", "HCM", "ICM", "NICM")
hsp.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A")
```


### load data

1. Human: the number of samples (n=197)

  * NC:  69
  
  * DCM: 63
  
  * ICM: 65

2. Mouse: the number of samples (n=85)

  * TAC_NC: 20 
    
  * TAC: 22 
    
  * HF_NC: 18
  
  * HF: 25
  
```{r}
hsp_DCM <- read.csv("../../Result/Differential/final/Hsp_DCM_NC_Diffgene.csv")
hsp_ICM <- read.csv("../../Result/Differential/final/Hsp_ICM_NC_Diffgene.csv")

mus_TAC <- read.csv("../../Result/Differential/final/Mus_TAC_NC_Diffgene.csv")
mus_HF <- read.csv("../../Result/Differential/final/Mus_HF_NC_Diffgene.csv")
```


### Curation Function 
```{r}
get_ORA <- function(dataset=hsp_DCM){
  
  # dataset=hsp_DCM
  
  # extract DEGs
  dat <- dataset %>% 
    dplyr::rename(SYMBOL=GeneID, log2FoldChange=logFC) %>%
    filter(Enrichment != "Nonsignif") %>%
    distinct() %>%
    arrange(log2FoldChange)
  
  # transform geneid
  require(org.Hs.eg.db)
  symbol2entrz <- AnnotationDbi::select(org.Hs.eg.db,
                                      key=dat$SYMBOL, 
                                      columns="ENTREZID",
                                      keytype="SYMBOL")
  symbol2entrz <- as_tibble(symbol2entrz)

  dat_v2 <- inner_join(symbol2entrz, dat, by ="SYMBOL") %>%
    dplyr::select(ENTREZID, log2FoldChange) %>%
    na.omit() %>% 
    distinct() %>% 
    group_by(ENTREZID) %>% 
    summarize(log2FoldChange=mean(log2FoldChange)) %>%
    ungroup() %>%
    arrange(desc(log2FoldChange))
  
  geneList <- tibble::deframe(dat_v2)  
  es_res <- enrichGO(gene          = names(geneList),
                     OrgDb         = org.Hs.eg.db,
                     keyType       = "ENTREZID",
                     ont           = "BP",
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.01,
                     qvalueCutoff  = 0.05)

  return(es_res)
}

plot_Escore <- function(dataset=GSEA_kegg,
                        number=10,
                        color=hsp.grp.col[2]){
  # dataset=es_res
  # number=10
  # color=hsp.grp.col[2]  
  
  edox <- setReadable(dataset, 'org.Hs.eg.db', 'ENTREZID')
  
  dat_es <- edox@result %>%
    filter(p.adjust < 0.01) %>%
    arrange(p.adjust)
  dat_plot <- dat_es[1:number, ] %>%
    mutate(value=-log10(p.adjust)) %>%
    arrange(value)
  dat_plot$Description <- factor(dat_plot$Description, levels = dat_plot$Description)
  
  pl <- ggplot(data = dat_plot, aes(x=Description, y=value))+
    geom_bar(stat = "identity", fill=color, color="black", width = 0.6)+
    labs(x="", y=expression(-log[10]("adjusted p-value")))+
    scale_y_continuous(trans = "log1p")+
    theme_classic()+
    coord_flip()+
    theme(axis.title = element_text(face = 'bold',color = 'black',size = 14),
                  axis.text = element_text(color = 'black',size = 10),
                  text = element_text(size = 8, color = "black", family="serif"),
                  legend.position = 'right',
                  legend.key.height = unit(0.6,'cm'),
                  legend.text = element_text(face = "bold", color = 'black',size = 10),
                  strip.text = element_text(face = "bold", size = 14))

  return(pl)
}
```


### Human DCM vs ICM
```{r}
DEGs_ORA_GO_DCM_NC <- get_ORA(dataset = hsp_DCM)
DEGs_ORA_GO_ICM_NC <- get_ORA(dataset = hsp_ICM)

DEGs_ORA_GO_barplot_DCM <- plot_Escore(dataset = DEGs_ORA_GO_DCM_NC, number = 10, color=hsp.grp.col[2])

DEGs_ORA_GO_barplot_ICM <- plot_Escore(dataset = DEGs_ORA_GO_ICM_NC, number = 10, color=hsp.grp.col[4])

DEGs_ORA_GO_DCM_ICM <- (DEGs_ORA_GO_barplot_DCM)/(DEGs_ORA_GO_barplot_ICM)+
  plot_layout(ncol = 1)
DEGs_ORA_GO_DCM_ICM
ggsave(filename = "../../Result/Figures/version_210322/Fig1D.pdf", 
       DEGs_ORA_GO_DCM_ICM, width = 7, height = 6)
```


### Mouse TAC vs HF
```{r}
DEGs_ORA_GO_TAC_NC <- get_ORA(dataset = mus_TAC)
DEGs_ORA_GO_HF_NC <- get_ORA(dataset = mus_HF)

DEGs_ORA_GO_barplot_TAC <- plot_Escore(dataset = DEGs_ORA_GO_TAC_NC, number = 10, color=mus.grp.col[2])

DEGs_ORA_GO_barplot_HF <- plot_Escore(dataset = DEGs_ORA_GO_HF_NC, number = 10, color=mus.grp.col[4])

DEGs_ORA_GO_TAC_HF <- (DEGs_ORA_GO_barplot_TAC)/(DEGs_ORA_GO_barplot_HF)+
  plot_layout(ncol = 1)
DEGs_ORA_GO_TAC_HF
ggsave(filename = "../../Result/Figures/version_210322/Fig1F.pdf", 
       DEGs_ORA_GO_TAC_HF, width = 7, height = 6)
```


### version
```{r}
sessionInfo()
```


### Reference 

1. [Functional Enrichment Analysis Methods](http://yulab-smu.top/clusterProfiler-book/chapter2.html) 
