---
title: "Curating Counts profile"
date: "Created: 2021-01-18 Updated: `r Sys.Date()`"
output: 
  html_notebook:
    codes: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(stringr)
library(data.table)
library(tibble)
library(ggplot2)

rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

colors <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A",
             "#993116", "#6798CE", "#146666", "#CE9924", "#6D659D",
             "#9F9B27", "#6D659D", "#9F9B27", "#C80b8A", "#2C3A89")
```


## Introduction

In this section, we convert ensembl id into gene symbol, transfor the musculus ensembl id into human gene symbol.


### load data
```{r}
hsp.set <- readRDS("../../Result/Profile/final/Homo_sapiens_ExprSet_VoomSNM.RDS")
mus.set <- readRDS("../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM.RDS")

mouse_gene <- fread("../../Result/GeneID/mouse_gene_all.tsv")
human_gene <- fread("../../Result/GeneID/human_gene_all.tsv")

mouse2human <- fread("../../Result/GeneID/gene_mouse2human.tsv")

phen <- read.csv("../../Result/Phenotype/phenotype_final.csv")
```


### convert ensemble into symbol id 

There are some duplicated gene symbols when converting the ensembl id into gene symbol because some gene symbols maybe have the same ensembl id, so we need to deduplicating. Only remain the highest expressed gene symbol via **median value**. Finally, restoring results as expressionset object to save.
```{r}
ensemble2symbol <- function(dataset=mus.set,
                            kind="mouse"){
  
  # dataset=mus.set
  # kind="mouse"
  
  pheno <- pData(dataset) %>% data.frame()
  edata <- exprs(dataset) %>% data.frame()
  
  if(kind == "mouse"){
    # change musculus ensemble id into gene symbol
    mdat <- inner_join(
      edata %>% rownames_to_column("ensembleid"), 
      mouse_gene %>% dplyr::select(ensembl_gene_id, external_gene_name), 
      by = c("ensembleid" = "ensembl_gene_id"))
  }else{
    # change human ensemble id into gene symbol
    mdat <- inner_join(
      edata %>% rownames_to_column("ensembleid"), 
      human_gene %>% dplyr::select(ensembl_gene_id, external_gene_name), 
      by = c("ensembleid" = "ensembl_gene_id"))
  }
    
  idx <- grep("ensembleid|external_gene_name", colnames(mdat))
    
  # filter features according gene symbol
  mdat$median <- apply(mdat[, -idx], 1, median)
  mdat <- with(mdat, mdat[order(external_gene_name, median, decreasing = T), ]) 
  mdat.deduplicated <- mdat[!duplicated(mdat$external_gene_name), ] %>% dplyr::select(-median) 
    
  # filter features according ensemble id
  mdat.deduplicated$median <- apply(mdat.deduplicated[, -idx], 1, median)
  mdat.deduplicated_v2 <- with(mdat.deduplicated, mdat.deduplicated[order(ensembleid, median, decreasing = T), ])
  mdat.deduplicated_v2 <- mdat.deduplicated_v2[!duplicated(mdat.deduplicated_v2$ensembleid), ] %>%
      dplyr::select(-median)
  rownames(mdat.deduplicated_v2) <- NULL
    
  # return result: gene symbol
  prof <- mdat.deduplicated_v2  %>% 
      dplyr::select(ensembleid, external_gene_name, everything()) %>%
      dplyr::rename(gene_symbol=external_gene_name) %>%
      dplyr::select(-ensembleid) %>%
      column_to_rownames("gene_symbol")
  
  
  # determine the right order between profile and phenotype 
  for(i in 1:ncol(prof)){ 
    if (!(colnames(prof)[i] == rownames(pheno)[i])) {
      stop(paste0(i, " Wrong"))
    }
  }  
  
  print(dim(prof))
  exprs <- as.matrix(prof)
  adf <-  new("AnnotatedDataFrame", data=pheno)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Convert ensembl id into gene symbol"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)
}


mouse2human_symbol <- function(dataset=mus.set){
  
  # dataset=mus.set
  
  pheno <- pData(dataset) %>% data.frame()
  edata <- exprs(dataset) %>% data.frame()
  
  # change musculus gene symbol into human gene symbol
  mdat <- inner_join(
      edata %>% rownames_to_column("ensembleid"), 
      mouse2human %>% dplyr::select(ensembl_id_mouse, HGNC_symbol), 
      by = c("ensembleid" = "ensembl_id_mouse"))
    
  idx <- grep("ensembleid|HGNC_symbol", colnames(mdat))
    
  # filter features according gene symbol
  mdat$median <- apply(mdat[, -idx], 1, median)
  mdat <- with(mdat, mdat[order(HGNC_symbol, median, decreasing = T), ]) 
  mdat.deduplicated <- mdat[!duplicated(mdat$HGNC_symbol), ] %>% dplyr::select(-median) 
    
  # filter features according ensemble id
  mdat.deduplicated$median <- apply(mdat.deduplicated[, -idx], 1, median)
  mdat.deduplicated_v2 <- with(mdat.deduplicated, mdat.deduplicated[order(ensembleid, median, decreasing = T), ])
  mdat.deduplicated_v2 <- mdat.deduplicated_v2[!duplicated(mdat.deduplicated_v2$ensembleid), ] %>%
      dplyr::select(-median)
  rownames(mdat.deduplicated_v2) <- NULL
    
  # return result: gene symbol
  prof <- mdat.deduplicated_v2  %>% 
      dplyr::select(ensembleid, HGNC_symbol, everything()) %>%
      dplyr::rename(gene_symbol=HGNC_symbol) %>%
      dplyr::select(-ensembleid) %>%
      column_to_rownames("gene_symbol")
  
  # determine the right order between profile and phenotype 
  for(i in 1:ncol(prof)){ 
    if (!(colnames(prof)[i] == rownames(pheno)[i])) {
      stop(paste0(i, " Wrong"))
    }
  }  
  
  print(dim(prof))
  exprs <- as.matrix(prof)
  adf <-  new("AnnotatedDataFrame", data=pheno)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Convert ensembl id into human gene symbol"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)

}


# visualization
pca_fun <- function(expers_set=hsp.set){
  
  # expers_set=hsp.set
  
  pheno <- pData(expers_set)
  edata <- exprs(expers_set)
  pca <- prcomp(t(edata), scale. = TRUE, center = T)
  score <- inner_join(pca$x %>% data.frame() %>% 
                        rownames_to_column("SampleID")  %>%  
                        dplyr::select(c(1:3)) , 
                        pheno %>% rownames_to_column("SampleID"),
                      by = "SampleID")

  pl <- ggplot(score, aes(x=PC1, y=PC2))+
          geom_point(aes(color=Batch), size=3.5)+
          guides(color=guide_legend(title=NULL,keywidth=.7, keyheight=.7))+
          scale_color_manual(values = colors)+
          theme_bw()+
          theme(axis.title = element_text(size=10, color="black", face="bold"),
                axis.text = element_text(size=9, color="black"),
                text = element_text(size=8, color="black", family="serif"),
                strip.text = element_text(size=9, color="black", face="bold", family="serif"), 
                panel.grid = element_blank(),
                legend.text=element_text(size=11, color = "black", family="serif"),
                legend.background = element_rect(color = "black", fill="white"))
  return(pl)  
}
```


### Homo sapiens profiles

* calculating
```{r}
hsp.set.cln <- ensemble2symbol(dataset = hsp.set, kind="human")
```

* plotting 
```{r, fig.width=14, fig.height=6}
if(!dir.exists("../../Result/Profile/final/")){
  dir.create("../../Result/Profile/final/")
}

cowplot::plot_grid(pca_fun(expers_set = hsp.set), 
                   pca_fun(expers_set = hsp.set.cln), 
                   ncol = 2,
                   labels = c("VoomSNM", "Convert"))

saveRDS(hsp.set.cln, "../../Result/Profile/final/Homo_sapiens_ExprSet_VoomSNM_symbol.RDS", compress = TRUE)
```


### Mus musculus profiles

* calculating
```{r}
mus.set.cln <- ensemble2symbol(dataset = mus.set, kind="mouse")

mus.set.HGNC <- mouse2human_symbol(dataset = mus.set)
```

* plotting 
```{r, fig.width=18, fig.height=6}
if(!dir.exists("../../Result/Profile/final/")){
  dir.create("../../Result/Profile/final/")
}

cowplot::plot_grid(pca_fun(expers_set = mus.set), 
                   pca_fun(expers_set = mus.set.cln),
                   pca_fun(expers_set = mus.set.HGNC),
                   ncol = 3,
                   labels = c("VoomSNM", "Convert", "HGNC"))
```

* split data into public and self-produced data 
```{r}
split_data_fun <- function(expers_set=mus.set.cln,
                           tag="self"){
  
  # expers_set=mus.set.cln
  # tag="self"
  
  pheno <- pData(expers_set)
  edata <- exprs(expers_set)
  
  if(tag == "self"){
    tag_list <- c("HF_b2", "HF_novo", "TAC_b1", "TAC_novo")
  }else{
    tag_list <- c("GSE133054_m", "GSE29446", "GSE66630")
  }

  datphe <- pheno %>% filter(Batch%in%tag_list) %>%
      rownames_to_column("SampleID") %>% 
      mutate(Batch=factor(as.character(Batch))) %>%
      column_to_rownames("SampleID")
  datprf <- edata %>% data.frame() %>% 
      dplyr::select(rownames(datphe)) %>%
      as.matrix()  
  print(dim(datprf))
  exprs <- datprf
  adf <-  new("AnnotatedDataFrame", data=datphe)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Convert ensembl id into human gene symbol"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)
  return(pl)  
}


mus.set.cln.self <- split_data_fun(expers_set = mus.set.cln, tag = "self")
mus.set.cln.pub <- split_data_fun(expers_set = mus.set.cln, tag = "public")
mus.set.HGNC.self <- split_data_fun(expers_set = mus.set.HGNC, tag = "self")
mus.set.HGNC.pub <- split_data_fun(expers_set = mus.set.HGNC, tag = "public")
```


* saving 
```{r}
saveRDS(mus.set.cln, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Msymbol.RDS", compress = TRUE)
saveRDS(mus.set.HGNC, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Hsymbol.RDS", compress = TRUE)

saveRDS(mus.set.cln.self, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Msymbol_self.RDS", compress = TRUE)
saveRDS(mus.set.cln.pub, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Msymbol_pub.RDS", compress = TRUE)

saveRDS(mus.set.HGNC.self, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Hsymbol_self.RDS", compress = TRUE)
saveRDS(mus.set.HGNC.pub, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Hsymbol_pub.RDS", compress = TRUE)
```


### 20210322: Integration of all the Data and show the correlations
```{r}
Integrated_matrix <- function(dataset1=hsp.set.cln,
                              dataset2=mus.set.HGNC,
                              metadata=phen){
  
  # dataset1=hsp.set.cln
  # dataset2=mus.set.HGNC
  # metadata=phen
  
  edata_1 <- exprs(dataset1) %>% data.frame()
  edata_2 <- exprs(dataset2) %>% data.frame()
  
  mdat <- inner_join(
      edata_1 %>% rownames_to_column("geneid"),
      edata_2 %>% rownames_to_column("geneid"),
      by = "geneid")
  phe <- metadata #%>%
    # filter(Group %in% c("NC","NICM","ICM","DCM",
    #                     "TAC_NC","TAC","HF_NC","HF","HCM"))
  
  idx <- intersect(phe$SampleID, colnames(mdat))
  pheno <- phe %>% filter(SampleID%in%idx) %>%
    column_to_rownames("SampleID")
    
  # return result: gene symbol
  prof <- mdat  %>% 
      dplyr::select(geneid, idx) %>%
      column_to_rownames("geneid")
  
  # determine the right order between profile and phenotype 
  for(i in 1:ncol(prof)){ 
    if (!(colnames(prof)[i] == rownames(pheno)[i])) {
      stop(paste0(i, " Wrong"))
    }
  }  
  
  print(dim(prof))
  exprs <- as.matrix(prof)
  adf <-  new("AnnotatedDataFrame", data=pheno)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Integrating Human and Mouse Expression Data"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)

}
Overall_set <- Integrated_matrix(dataset1=hsp.set.cln,
                  dataset2=mus.set.HGNC,
                  metadata=phen)

saveRDS(Overall_set, 
        "../../Result/Profile/final/Hsp_Mus_ExprSet_VoomSNM.RDS", compress = TRUE)
```


### version
```{r}
sessionInfo()
```


### Reference 

* [biomaRt tutorial](https://www.bioconductor.org/packages/devel/bioc/vignettes/biomaRt/inst/doc/biomaRt.html)

* [gene id convert](https://cloud.tencent.com/developer/news/324582) 