---
title: "Enriched pathway analysis: based on cluterProfiler and enrichplot"
date: "Created: 2021-02-03 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

## Introduction

Performing Enriched pathway analysis via **clusterProfiler** and **enrichplot** package based on the DESeq2 results

* *clusterProfiler* calculates the enrichment score

* *enrichplot* displays the enrichment result

Two approaches for enrichment analysis:

* Over Representation Analysis: the up- or down-regulated DEGs 

* Gene Set Enrichment Analysis: All the ranked DEGs

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(ggplot2)
library(data.table)
library(clusterProfiler)
library(enrichplot)
library(patchwork)

rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

mus.grp <- c("TAC_NC", "TAC", "HF_NC", "HF")
mus.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F")

hsp.grp <- c("NC", "DCM", "HCM", "ICM", "NICM")
hsp.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A")
```


### load data 

* gene Expression DESeq2 object
```{r}
DCM_dif <- fread("../../Result/Differential/Hsp_DCM_NC_Diffgene.csv")
ICM_dif <- fread("../../Result/Differential/Hsp_ICM_NC_Diffgene.csv")
HF_dif <- fread("../../Result/Differential/Mus_HF_NC_Diffgene.csv")
TAC_dif <- fread("../../Result/Differential/Mus_TAC_NC_Diffgene.csv")
```


### Curation Function 

* Over Representation Analysis: the DEgenes vectors (up or down)

Questions: Whether the genes of Mouse are suitable for this analysis
```{r}
# ORA 
get_ORA <- function(dataset=DCM_dif,
                    group=hsp.grp[2]){
  
  # dataset=DCM_dif
  # group=hsp.grp[2]
  
  # extract DEGs
  dat <- dataset %>% 
    dplyr::rename(SYMBOL=GeneID, log2FoldChange=logFC) %>%
    filter(Enrichment%in%group) %>%
    distinct() %>%
    arrange(log2FoldChange)
  
  # transform geneid
  require(org.Hs.eg.db)
  symbol2entrz <- AnnotationDbi::select(org.Hs.eg.db,
                                      key=dat$SYMBOL, 
                                      columns="ENTREZID",
                                      keytype="SYMBOL")
  symbol2entrz <- as_tibble(symbol2entrz)

  dat_v2 <- inner_join(symbol2entrz, dat, by ="SYMBOL") %>%
    dplyr::select(ENTREZID, log2FoldChange) %>%
    na.omit() %>% 
    distinct() %>% 
    group_by(ENTREZID) %>% 
    summarize(log2FoldChange=mean(log2FoldChange)) %>%
    ungroup() %>%
    arrange(desc(log2FoldChange))
  
  geneList <- tibble::deframe(dat_v2)  
  es_res <- enrichGO(gene          = names(geneList),
                     OrgDb         = org.Hs.eg.db,
                     keyType       = "ENTREZID",
                     ont           = "CC",
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.01,
                     qvalueCutoff  = 0.05)
  
  res <- list(glist=geneList, es=es_res) 
  
  return(res)
}

plot_Escore <- function(dataset=GSEA_kegg,
                        type="barplot",
                        number=10){
  # dataset=GSEA_kegg
  # type="barplot"
  # number=10
  
  my_theme <- theme(axis.title = element_text(face = 'bold',color = 'black',size = 14),
                  axis.text = element_text(color = 'black',size = 10),
                  text = element_text(size = 8, color = "black", family="serif"),
                  legend.position = 'right',
                  legend.key.height = unit(0.6,'cm'),
                  legend.text = element_text(face = "bold", color = 'black',size = 10),
                  strip.text = element_text(face = "bold", size = 14))
  
  edox <- setReadable(dataset$es, 'org.Hs.eg.db', 'ENTREZID')
  
  if(type == "barplot"){
    pl <- barplot(edox, showCategory = number)+
      my_theme
  }else if(type == "dotplot"){
    pl <- enrichplot::dotplot(edox, showCategory=number)+
      my_theme
  }else if(type == "network"){
    pl <- enrichplot::cnetplot(edox, categorySize="pvalue", foldChange=dataset$glist, showCategory = 3)+
      my_theme
  }else if(type == "heatmap"){
    pl <- enrichplot::heatplot(edox, foldChange=dataset$glist, showCategory = number)+
      my_theme
  }else if(type == "enrichment"){
    pl <- enrichplot::emapplot(dataset$es, pie_scale=1.5, layout="kk", showCategory = number)+
      my_theme
  }else if(type == "UpSet"){
    require(ggupset)
    pl <- enrichplot::upsetplot(dataset$es, n=number)+
      my_theme
  }else if(type == "pubmed"){                    
    terms <- dataset$es$Description[1:number]
    pl <- enrichplot::pmcplot(terms, 2010:2021, proportion = T)+
      my_theme
  }
  
  return(pl)
}
```


### Over Representation Analysis

* Calculate the Enrichment results: DCM_NC
```{r}
DEGs_ORA_GO_DCM_NC <- get_ORA(dataset = DCM_dif, group = hsp.grp[c(1:2)])
DEGs_ORA_GO_NC <- get_ORA(dataset = DCM_dif, group = hsp.grp[1])
DEGs_ORA_GO_DCM <- get_ORA(dataset = DCM_dif, group = hsp.grp[2])

save(DEGs_ORA_GO_DCM_NC, DEGs_ORA_GO_NC, DEGs_ORA_GO_DCM,
     file = "../../Result/Enrichment/DEGs_ORA_GO_DCM_NC.RData")
```


* Visualization : GO barplot
```{r, fig.width=16, fig.height=8}
DEGs_ORA_GO_barplot_DCM_NC <- plot_Escore(dataset = DEGs_ORA_GO_DCM_NC, number = 10, type = "barplot")
DEGs_ORA_GO_barplot_DCM <- plot_Escore(dataset = DEGs_ORA_GO_DCM, number = 10, type = "barplot")

(DEGs_ORA_GO_barplot_DCM_NC | DEGs_ORA_GO_barplot_DCM)+
  plot_layout(guides = 'collect')+
  plot_annotation(tag_levels = "A")
```

* Calculate the Enrichment results: ICM_NC
```{r}
DEGs_ORA_GO_ICM_NC <- get_ORA(dataset = ICM_dif, group = hsp.grp[c(1, 4)])
DEGs_ORA_GO_ICM <- get_ORA(dataset = ICM_dif, group = hsp.grp[4])

save(DEGs_ORA_GO_ICM_NC, DEGs_ORA_GO_ICM,
     file = "../../Result/Enrichment/DEGs_ORA_GO_ICM_NC.RData")
```


* Visualization : GO barplot
```{r, fig.width=16, fig.height=8}
DEGs_ORA_GO_barplot_ICM_NC <- plot_Escore(dataset = DEGs_ORA_GO_ICM_NC, number = 10, type = "barplot")
DEGs_ORA_GO_barplot_ICM <- plot_Escore(dataset = DEGs_ORA_GO_ICM, number = 10, type = "barplot")

(DEGs_ORA_GO_barplot_ICM_NC | DEGs_ORA_GO_barplot_ICM)+
  plot_layout(guides = 'collect')+
  plot_annotation(tag_levels = "A")
```

### version
```{r}
sessionInfo()
```


### Reference 

1. [Functional Enrichment Analysis Methods](http://yulab-smu.top/clusterProfiler-book/chapter2.html) 

2. [Visualization of Functional Enrichment Result](http://yulab-smu.top/clusterProfiler-book/chapter12.html#enrichment-map)
