---
title: "Enriched pathway analysis: fgsea"
date: "Created: 2021-01-29 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

## Introduction

Performing Enriched pathway analysis via fgsea package

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(ggplot2)
library(data.table)
library(fgsea)

rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

mus.grp <- c("TAC_NC", "TAC", "HF_NC", "HF")
mus.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F")

hsp.grp <- c("NC", "DCM", "HCM", "ICM", "NICM")
hsp.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A")
```


### load data 

* gene Expression DESeq2 object
```{r}
DCM_NC_DEGs <- read.csv("../../Result/Differential/Hsp_DCM_NC_Diffgene.csv")
ICM_NC_DEGs <- read.csv("../../Result/Differential/Hsp_ICM_NC_Diffgene.csv")

TAC_NC_DEGs <- read.csv("../../Result/Differential/Mus_TAC_HF_Diffgene.csv")
HF_NC_DEGs <- read.csv("../../Result/Differential/Mus_HF_NC_Diffgene.csv")

pathways_hallmark_kegg <- gmtPathways("../../Result/GeneSetdb/msigdb.v7.1.symbols_KEGG.gmt")
pathways_hallmark_GO <- gmtPathways("../../Result/GeneSetdb/msigdb.v7.1.symbols_GO.gmt")
```


### Curation Function 

the Genes were ranked by the **stat** index.
```{r}
get_fgsea <- function(dataset=DCM_NC_DEGs,
                      pathways=pathways_hallmark_kegg){
  
  # dataset=DCM_NC_DEGs
  # pathways=pathways_hallmark_kegg
  
  dat <- dataset %>% data.frame() %>%
    dplyr::rename(SYMBOL=GeneID) %>%
    dplyr::select(SYMBOL, t) %>% 
    na.omit() %>% 
    distinct() %>% 
    group_by(SYMBOL) %>% 
    summarize(t=mean(t))
  
  ranks <- deframe(dat)

  fgseaRes <- fgsea(pathways=pathways, 
                    stats=ranks, 
                    nperm=1000)
  
  pathways_gene <- pathways %>% 
                    enframe("pathway", "SYMBOL") %>% 
                    tidyr::unnest(cols = c(SYMBOL)) %>% 
                    inner_join(dataset %>% dplyr::rename(SYMBOL=GeneID),
                               by="SYMBOL") 
  res <- list(ranks=ranks, fgres=fgseaRes, pgene=pathways_gene)
  
  return(res)
}

# barplot
get_plot <- function(dataset=DCM_NC_kegg_gsea,
                     pathways=pathways_hallmark_kegg,
                     number=10,
                     tag="kegg",
                     group_name=c("DCM", "NC"),
                     group_col=c("#77A2D1", "#6C326C")){

  # dataset=DCM_NC_kegg_gsea
  # pathways=pathways_hallmark_kegg
  # number=10
  # tag="kegg"
  # group_name=c("DCM", "NC")
  # group_col=c("#77A2D1", "#6C326C")  
  
  fgseaRes <- dataset$fgres
  ranks <- dataset$ranks

  topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=number), pathway]
  topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=number), pathway]
  topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
  
  if(tag == "kegg"){
    names <- paste0("../../Result/Enrichment/KEGG_topPathway_", number, ".pdf")    
  }else{
    names <- paste0("../../Result/Enrichment/GO_topPathway_", number, ".pdf")    
  }

  pdf(file = names, width = 11, height = 10)
  plotGseaTable(pathways[topPathways], 
                ranks, 
                fgseaRes, 
                gseaParam=0.5)
  dev.off()
  
  
  fgseaResTidy <- fgseaRes %>%
    as_tibble() %>%
    filter(pathway%in%topPathways) %>%
    arrange(desc(NES))%>%
    mutate(color=ifelse(NES < 0, group_name[1], group_name[2]),
           color=factor(color, levels = group_name))
  
 pl_bar <- ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES))+
    geom_col(aes(fill = color))+
    coord_flip()+
    scale_fill_manual(values = group_col)+
    labs(x="Pathway", y="Normalized Enrichment Score",
         title="Hallmark pathways NES from GSEA") + 
    guides(fill=guide_legend("Cluster"))+
    theme_minimal()+
    theme(axis.title = element_text(face = 'bold',color = 'black',size = 14),
                  axis.text = element_text(color = 'black',size = 10),
                  text = element_text(size = 8, color = "black", family="serif"),
                  legend.position = 'right',
                  legend.key.height = unit(0.6,'cm'),
                  legend.text = element_text(face = "bold", color = 'black',size = 10),
                  strip.text = element_text(face = "bold", size = 14))
 
 return(pl_bar)
}

# es per pathway 
get_es_plot <- function(dataset=kegg_gsea,
                        pathways=pathways_hallmark_kegg,
                        pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON"){
  
  # dataset=DCM_NC_kegg_gsea
  # pathways=pathways_hallmark_kegg
  # pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON"
  
  
  pl <- plotEnrichment(pathways[[pathway_name]], dataset$ranks)+
    labs(title=pathway_name,
         x="Rank in Ordered DataSet",
         y="Enrichment Score(ES)")+
    theme(axis.title = element_text(face = 'bold',color = 'black',size = 14),
                  axis.text = element_text(color = 'black',size = 10),
                  text = element_text(size = 8, color = "black", family="serif"),
                  legend.position = 'right',
                  legend.key.height = unit(0.6,'cm'),
                  legend.text = element_text(face = "bold", color = 'black',size = 10),
                  strip.text = element_text(face = "bold", size = 14)) 
  return(pl)
}
```


### KEGG 

* DCM vs NC 
```{r}
DCM_NC_kegg_gsea <- get_fgsea(dataset = DCM_NC_DEGs, pathways = pathways_hallmark_kegg)

if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}
saveRDS(DCM_NC_kegg_gsea, "../../Result/Enrichment/DCM_NC_KEGG_pathway_gsea.RDS")
write.csv(DCM_NC_kegg_gsea$pgene, "../../Result/Enrichment/DCM_NC_KEGG_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
DCM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl <- 
  get_es_plot(dataset=DCM_NC_kegg_gsea, pathways=pathways_hallmark_kegg,
         pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON")
ggsave("../../Result/Enrichment/DCM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON.pdf",
       DCM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl, width = 5, height = 4, dpi = 600)

# top pathway
DCM_NC_kegg_bar <- get_plot(dataset=DCM_NC_kegg_gsea, pathways=pathways_hallmark_kegg, 
                     number=10, tag="kegg", 
                     group_name=c("DCM", "NC"), 
                     group_col=c("#77A2D1", "#6C326C"))
DCM_NC_kegg_bar
ggsave("../../Result/Enrichment/DCM_NC_KEGG_barplot.pdf", DCM_NC_kegg_bar, width = 10, height = 8, dpi = 600)
```


* ICM vs NC 
```{r}
ICM_NC_kegg_gsea <- get_fgsea(dataset = ICM_NC_DEGs, pathways = pathways_hallmark_kegg)

if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}
saveRDS(ICM_NC_kegg_gsea, "../../Result/Enrichment/ICM_NC_KEGG_pathway_gsea.RDS")
write.csv(ICM_NC_kegg_gsea$pgene, "../../Result/Enrichment/ICM_NC_KEGG_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
ICM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl <- 
  get_es_plot(dataset=ICM_NC_kegg_gsea, pathways=pathways_hallmark_kegg,
         pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON")
ggsave("../../Result/Enrichment/ICM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON.pdf",
       ICM_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl, width = 5, height = 4, dpi = 600)

# top pathway
ICM_NC_kegg_bar <- get_plot(dataset=ICM_NC_kegg_gsea, pathways=pathways_hallmark_kegg, 
                     number=10, tag="kegg", 
                     group_name=c("ICM", "NC"), 
                     group_col=c("#635F5F", "#6C326C"))
ICM_NC_kegg_bar
ggsave("../../Result/Enrichment/ICM_NC_KEGG_barplot.pdf", ICM_NC_kegg_bar, width = 10, height = 8, dpi = 600)
```

* TAC vs TAC_NC 
```{r}
TAC_NC_kegg_gsea <- get_fgsea(dataset = TAC_NC_DEGs, pathways = pathways_hallmark_kegg)

if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}
saveRDS(TAC_NC_kegg_gsea, "../../Result/Enrichment/TAC_NC_KEGG_pathway_gsea.RDS")
write.csv(TAC_NC_kegg_gsea$pgene, "../../Result/Enrichment/TAC_NC_KEGG_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
TAC_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl <- 
  get_es_plot(dataset=TAC_NC_kegg_gsea, pathways=pathways_hallmark_kegg,
         pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON")
ggsave("../../Result/Enrichment/TAC_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON.pdf",
       TAC_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl, width = 5, height = 4, dpi = 600)

# top pathway
TAC_NC_kegg_bar <- get_plot(dataset=TAC_NC_kegg_gsea, pathways=pathways_hallmark_kegg, 
                     number=10, tag="kegg", 
                     group_name=c("TAC", "TAC_NC"), 
                     group_col=c("#77A2D1", "#6C326C"))
TAC_NC_kegg_bar
ggsave("../../Result/Enrichment/TAC_NC_KEGG_barplot.pdf", TAC_NC_kegg_bar, width = 10, height = 8, dpi = 600)
```


* HF vs HF_NC 
```{r}
HF_NC_kegg_gsea <- get_fgsea(dataset = HF_NC_DEGs, pathways = pathways_hallmark_kegg)

if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}
saveRDS(HF_NC_kegg_gsea, "../../Result/Enrichment/HF_NC_KEGG_pathway_gsea.RDS")
write.csv(HF_NC_kegg_gsea$pgene, "../../Result/Enrichment/HF_NC_KEGG_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
HF_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl <- 
  get_es_plot(dataset=HF_NC_kegg_gsea, pathways=pathways_hallmark_kegg,
         pathway_name="KEGG_REGULATION_OF_ACTIN_CYTOSKELETON")
ggsave("../../Result/Enrichment/HF_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON.pdf",
       HF_NC_KEGG_REGULATION_OF_ACTIN_CYTOSKELETON_pl, width = 5, height = 4, dpi = 600)

# top pathway
HF_NC_kegg_bar <- get_plot(dataset=HF_NC_kegg_gsea, pathways=pathways_hallmark_kegg, 
                     number=10, tag="kegg", 
                     group_name=c("HF", "HF_NC"), 
                     group_col=c("#635F5F", "#FFD169"))
HF_NC_kegg_bar
ggsave("../../Result/Enrichment/HF_NC_KEGG_barplot.pdf", HF_NC_kegg_bar, width = 10, height = 8, dpi = 600)
```

### GO 

* DCM vs NC 
```{r, fig.width=15}
if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}

DCM_NC_GO_gsea <- get_fgsea(dataset = DCM_NC_DEGs, pathways = pathways_hallmark_GO)
saveRDS(DCM_NC_GO_gsea, "../../Result/Enrichment/DCM_NC_GO_pathway_gsea.RDS")
write.csv(DCM_NC_GO_gsea$pgene, "../../Result/Enrichment/DCM_NC_GO_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
DCM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl <- 
  get_es_plot(dataset=DCM_NC_GO_gsea, pathways=pathways_hallmark_GO,
         pathway_name="GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION")
DCM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl
ggsave("../../Result/Enrichment/DCM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION.pdf",
       DCM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl, width = 5, height = 4, dpi = 600)

# top pathway
DCM_NC_GO_bar <- get_plot(dataset=DCM_NC_GO_gsea, 
                          pathways=pathways_hallmark_GO, 
                          number=10, tag="GO", 
                          group_name=c("DCM", "NC"), 
                          group_col=c("#77A2D1", "#6C326C"))
DCM_NC_GO_bar
ggsave("../../Result/Enrichment/DCM_NC_GO_barplot.pdf", DCM_NC_GO_bar, width = 15, height = 8, dpi = 600)
```

* ICM vs NC 
```{r, fig.width=15}
if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}

ICM_NC_GO_gsea <- get_fgsea(dataset = ICM_NC_DEGs, pathways = pathways_hallmark_GO)
saveRDS(ICM_NC_GO_gsea, "../../Result/Enrichment/ICM_NC_GO_pathway_gsea.RDS")
write.csv(ICM_NC_GO_gsea$pgene, "../../Result/Enrichment/ICM_NC_GO_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
ICM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl <- 
  get_es_plot(dataset=ICM_NC_GO_gsea, pathways=pathways_hallmark_GO,
         pathway_name="GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION")
ICM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl
ggsave("../../Result/Enrichment/ICM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION.pdf",
       ICM_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl, width = 5, height = 4, dpi = 600)

# top pathway
ICM_NC_GO_bar <- get_plot(dataset=ICM_NC_GO_gsea, 
                          pathways=pathways_hallmark_GO, 
                          number=10, tag="GO", 
                          group_name=c("ICM", "NC"), 
                          group_col=c("#FFD169", "#6C326C"))
ICM_NC_GO_bar
ggsave("../../Result/Enrichment/ICM_NC_GO_barplot.pdf", ICM_NC_GO_bar, width = 15, height = 8, dpi = 600)
```

* TAC vs TAC_NC
```{r, fig.width=15}
if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}

TAC_NC_GO_gsea <- get_fgsea(dataset = TAC_NC_DEGs, pathways = pathways_hallmark_GO)
saveRDS(TAC_NC_GO_gsea, "../../Result/Enrichment/TAC_NC_GO_pathway_gsea.RDS")
write.csv(TAC_NC_GO_gsea$pgene, "../../Result/Enrichment/TAC_NC_GO_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
TAC_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl <- 
  get_es_plot(dataset=TAC_NC_GO_gsea, pathways=pathways_hallmark_GO,
         pathway_name="GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION")
TAC_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl
ggsave("../../Result/Enrichment/TAC_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION.pdf",
       TAC_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl, width = 5, height = 4, dpi = 600)

# top pathway
TAC_NC_GO_bar <- get_plot(dataset=TAC_NC_GO_gsea, 
                          pathways=pathways_hallmark_GO, 
                          number=10, tag="GO", 
                          group_name=c("TAC", "TAC_NC"), 
                          group_col=c("#77A2D1", "#6C326C"))
TAC_NC_GO_bar
ggsave("../../Result/Enrichment/TAC_NC_GO_barplot.pdf", TAC_NC_GO_bar, width = 15, height = 8, dpi = 600)
```

* HF vs HF_NC 
```{r, fig.width=15}
if(!dir.exists("../../Result/Enrichment")){
  dir.create("../../Result/Enrichment", recursive = T)
}

HF_NC_GO_gsea <- get_fgsea(dataset = HF_NC_DEGs, pathways = pathways_hallmark_GO)
saveRDS(HF_NC_GO_gsea, "../../Result/Enrichment/HF_NC_GO_pathway_gsea.RDS")
write.csv(HF_NC_GO_gsea$pgene, "../../Result/Enrichment/HF_NC_GO_pathway_gsea_gene.csv", row.names = F)

# plot per pathway 
HF_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl <- 
  get_es_plot(dataset=HF_NC_GO_gsea, pathways=pathways_hallmark_GO,
         pathway_name="GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION")
HF_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl
ggsave("../../Result/Enrichment/HF_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION.pdf",
       HF_NC_GO_POSITIVE_REGULATION_OF_VIRAL_TRANSCRIPTION_pl, width = 5, height = 4, dpi = 600)

# top pathway
HF_NC_GO_bar <- get_plot(dataset=HF_NC_GO_gsea, 
                          pathways=pathways_hallmark_GO, 
                          number=10, tag="GO", 
                          group_name=c("HF", "HF_NC"), 
                          group_col=c("#635F5F", "#FFD169"))
HF_NC_GO_bar
ggsave("../../Result/Enrichment/HF_NC_GO_barplot.pdf", HF_NC_GO_bar, width = 15, height = 8, dpi = 600)
```


### version
```{r}
sessionInfo()
```


### Reference 

1. [Fast Gene Set Enrichment Analysis](https://bioconductor.org/packages/release/bioc/manuals/fgsea/man/fgsea.pdf) 

2. [DESeq results to pathways in 60 Seconds with the fgsea package](https://stephenturner.github.io/deseq-to-fgsea/)

