---
title: "The Overall Correlation between groups: from raw counts to transformed data"
date: "Created: 2021-03-30 Updated: `r Sys.Date()`"
output: 
  html_notebook:
    codes: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(data.table)
library(tibble)
library(convert)
library(ggplot2)

library(limma)
library(edgeR)
library(snm)
library(gbm)


# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

mus.grp <- c("TAC_NC", "TAC", "HF_NC", "HF")
mus.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F")

hsp.grp <- c("NC", "DCM", "HCM", "ICM", "NICM")
hsp.grp.col <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A")

colors <- c("#6C326C", "#77A2D1", "#FFD169", "#635F5F", "#D4D09A",
            "#993116", "#6798CE", "#146666", "#CE9924","#9F9B27", 
            "#6D659D", "#C80b8A", "#2C3A89", "#C8C5C5", "#90E2BF", 
            "#FDAB4D", "#F4F4E8", "#B054BF", "#FCE873", "#FFCCDB", 
            "#AFD300", "#B089D8", "#F96E6F", "#AAD3ED", "#639BCE")
```


### load data 
```{r}
# human count profile
hsp_GSE116250 <- fread("../../Result/Profile/bulkRNA/hsp_GSE116250_stringtie_counts.csv")
hsp_GSE120852 <- fread("../../Result/Profile/bulkRNA/hsp_GSE120852_stringtie_counts.csv")
hsp_GSE133054 <- fread("../../Result/Profile/bulkRNA/hsp_GSE133054_stringtie_counts.csv")
hsp_GSE135055 <- fread("../../Result/Profile/bulkRNA/hsp_GSE135055_stringtie_counts.csv")
hsp_GSE147236 <- fread("../../Result/Profile/bulkRNA/hsp_GSE147236_stringtie_counts.csv")
hsp_GSE46224 <- fread("../../Result/Profile/bulkRNA/hsp_GSE46224_stringtie_counts.csv")
hsp_GSE48166 <- fread("../../Result/Profile/bulkRNA/hsp_GSE48166_stringtie_counts.csv")
hsp_GSE57344 <- fread("../../Result/Profile/bulkRNA/hsp_GSE57344_stringtie_counts.csv")

# musculus count profile
mus_GSE133054 <- fread("../../Result/Profile/bulkRNA/mus_GSE133054_stringtie_counts.csv")
mus_GSE29446 <- fread("../../Result/Profile/bulkRNA/mus_GSE29446_stringtie_counts.csv")
mus_GSE66630 <- fread("../../Result/Profile/bulkRNA/mus_GSE66630_stringtie_counts.csv")
mus_self <- fread("../../Result/Profile/bulkRNA/mus_self_stringtie_counts.csv")

# phenotype 
phen <- read.csv("../../Result/Phenotype/Heart_failure_phenotype_20210118.csv")
phen <- phen %>% mutate(Batch=gsub("_h|_m", "", Batch))

# gene annotation
mouse_gene <- fread("../../Result/GeneID/mouse_gene_all.tsv")
human_gene <- fread("../../Result/GeneID/human_gene_all.tsv")
mouse2human <- fread("../../Result/GeneID/gene_mouse2human.tsv")
```


### Merge and transform 

the following scripts comprise of several functions: filtering features, removing batch effects and visualizing  

  1. the transcript_biotype of gene descripted *protein_coding* would be remained

  2. the feature of occurrence less than 0.2
  
  3. the feature of total counts more than 10 
  
  4. the unique number of feature(Avoid the constant expression value per gene) 

```{r}
get_filter_prf <- function(x=phen, 
                           y1=hsp.prf.lst,
                           y2=mus.prf.lst,
                           uniq=2, 
                           occurrence=0.2,
                           ncount=10){
  # x=phen
  # y1=hsp.prf.lst
  # y2=mus.prf.lst
  # occurrence=0.2
  # uniq=2
  # ncount=10
  
  transform <- function(dataset=y1,
                        geneset=human_gene,
                        kind="human"){
    # dataset=y2
    # geneset=mouse_gene
    # kind="mouse"
  
    prof <- dataset[[1]]
    for(i in 2:length(dataset)){
      prof <- inner_join(prof, dataset[[i]], by="gene_id")
    }
    prof$gene_id <- gsub("\\|\\S+", "", prof$gene_id)
    
    gene_list <- geneset %>% filter(transcript_biotype == "protein_coding") 
    
    gene_sid <- intersect(gene_list$ensembl_gene_id, prof$gene_id)
    gene_list_filter <- gene_list %>% filter(ensembl_gene_id %in% gene_sid)
    prf <- prof %>% filter(gene_id %in% as.character(gene_sid)) %>% 
      column_to_rownames("gene_id")
    
    sid <- intersect(x$SampleID, colnames(prf))
    
    # phenotype
    phe <- x %>% filter(SampleID%in%sid) %>%
      mutate(Batch=factor(as.character(Batch)),
             Group=factor(as.character(Group))) %>%
      mutate(batch_number=as.numeric(Batch)) %>%
      column_to_rownames("SampleID") 
    
    # profile by occurrence
    prf_occ <- prf %>% rownames_to_column("tmp") %>% 
      filter(apply(dplyr::select(., -one_of("tmp")), 1, function(x) {
              sum(x != 0)/length(x)}) > occurrence) %>%
      dplyr::select(c(tmp, rownames(phe))) %>%
      column_to_rownames("tmp")
    
    # filter profile by unique  
    prf_uniq_number <- apply(prf_occ, 1, function(x){length(unique(x[x > 0]))}) %>% 
      data.frame() %>% setNames("number")
    remain_genes <- prf_uniq_number %>% filter(number > uniq)
    
    prf_count <- prf_occ[rowSums(prf_occ) > ncount, ]
    prf_final <- prf_count[rownames(prf_count)%in%rownames(remain_genes), ]
    
    # determine the right order between profile and phenotype 
    for(i in 1:ncol(prf_final)){ 
      if (!(colnames(prf_final)[i] == rownames(phe)[i])) {
        stop(paste0(i, " Wrong"))
      }
    }  
    
    if(kind == "human"){
        mdat <- inner_join(
          prf_final %>% rownames_to_column("ensembleid"), 
          human_gene %>% dplyr::select(ensembl_gene_id, external_gene_name), 
          by = c("ensembleid" = "ensembl_gene_id"))      
    }else{
        mdat <- inner_join(
            prf_final %>% rownames_to_column("ensembleid"), 
            mouse2human %>% dplyr::select(ensembl_id_mouse, HGNC_symbol), 
            by = c("ensembleid" = "ensembl_id_mouse")) %>%
          dplyr::rename(external_gene_name=HGNC_symbol)
    }

    idx <- grep("ensembleid|external_gene_name", colnames(mdat))
      
    # filter features according gene symbol
    mdat$median <- apply(mdat[, -idx], 1, median)
    mdat <- with(mdat, mdat[order(external_gene_name, median, decreasing = T), ]) 
    mdat.deduplicated <- mdat[!duplicated(mdat$external_gene_name), ] %>% dplyr::select(-median) 
      
    # filter features according ensemble id
    mdat.deduplicated$median <- apply(mdat.deduplicated[, -idx], 1, median)
    mdat.deduplicated_v2 <- with(mdat.deduplicated, 
                                 mdat.deduplicated[order(ensembleid, median, decreasing = T), ])
    mdat.deduplicated_v2 <- mdat.deduplicated_v2[!duplicated(mdat.deduplicated_v2$ensembleid), ] %>%
        dplyr::select(-median)
    rownames(mdat.deduplicated_v2) <- NULL
      
    # return result: gene symbol
    prof <- mdat.deduplicated_v2  %>% 
        dplyr::select(ensembleid, external_gene_name, everything()) %>%
        dplyr::rename(gene_symbol=external_gene_name) %>%
        dplyr::select(-ensembleid) %>%
        column_to_rownames("gene_symbol")
    
    # determine the right order between profile and phenotype 
    for(i in 1:ncol(prof)){ 
      if (!(colnames(prof)[i] == rownames(phe)[i])) {
        stop(paste0(i, " Wrong"))
      }
    }     
    
    res <- list(phe=phe, prf=prof)
    return(res)
  }
  
  hsp_set <- transform(dataset = y1, geneset = human_gene, kind="human")
  mus_set <- transform(dataset = y2, geneset = mouse_gene, kind="mouse")
  
  print(table(hsp_set$phe$Batch, hsp_set$phe$Group))
  print(table(mus_set$phe$Batch, mus_set$phe$Group))
  
  # merge data 
  phe_merge <- rbind(hsp_set$phe, mus_set$phe)
  prf_merge <- inner_join(hsp_set$prf %>% rownames_to_column("Symbol"),
                          mus_set$prf %>% rownames_to_column("Symbol"),
                          by = "Symbol") %>%
    column_to_rownames("Symbol")
  
  # determine the right order between profile and phenotype 
  for(i in 1:ncol(prf_merge)){ 
    if (!(colnames(prf_merge)[i] == rownames(phe_merge)[i])) {
      stop(paste0(i, " Wrong"))
    }
  }   
  
  exprs <- as.matrix(prf_merge)
  adf <-  new("AnnotatedDataFrame", data=phe_merge)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Merge hsp&mus profile"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)
}
```

```{r}
hsp.prf.lst <- list(hsp_GSE116250, hsp_GSE120852, hsp_GSE133054, hsp_GSE135055, 
                    hsp_GSE147236, hsp_GSE46224, hsp_GSE48166, hsp_GSE57344)
mus.prf.lst <- list(mus_GSE133054, mus_GSE29446, mus_GSE66630, mus_self)

Count_ExprSet <- get_filter_prf(x=phen, 
               y1=hsp.prf.lst,
               y2=mus.prf.lst,
               uniq=2, 
               occurrence=0.2,
               ncount=10)

#saveRDS(Count_ExprSet, "../../Result/Profile/final/hsp_mus_ExprSet_counts.RDS", compress = TRUE)
```


### Voom+SNM remove batch effects
```{r}
# Count_ExprSet <- readRDS("../../Result/Profile/final/hsp_mus_ExprSet_counts.RDS")

reduce_voom_SNM <- function(x=Count_ExprSet){
  
  # x=Count_ExprSet

  qcMetadata <- pData(x) # ADAPT THIS AS NEEDED
  qcData <- t(exprs(x))  # ADAPT THIS AS NEEDED
    
  # Set up design matrix
  covDesignNorm <- model.matrix(~0 + Batch + Group, 
                                      data = qcMetadata)
  
  # Check row dimensions
  dim(covDesignNorm)[1] == dim(qcData)[1]
    
  print(colnames(covDesignNorm))
  # The following corrects for column names that are incompatible with downstream processing
  colnames(covDesignNorm) <- gsub('([[:punct:]])|\\s+', '', colnames(covDesignNorm))
  print(colnames(covDesignNorm))
    
  # Set up counts matrix
  counts <- t(qcData) # DGEList object from a table of counts (rows=features, columns=samples)
    
  # Quantile normalize and plug into voom
  dge <- DGEList(counts = counts)
  vdge <- voom(dge, 
               design = covDesignNorm, 
               plot = TRUE, 
               save.plot = TRUE,
               normalize.method = "none")
    
  # List biological and normalization variables in model matrices
  bio.matrix <- model.matrix( ~ Group, data=qcMetadata)
  colnames(bio.matrix) <- gsub('([[:punct:]])|\\s+', '', colnames(bio.matrix))
  
  adj.matrix <- model.matrix( ~ Batch, data=qcMetadata)
  colnames(adj.matrix) <- gsub('([[:punct:]])|\\s+', '', colnames(adj.matrix))
  
  print(dim(adj.matrix))
  print(dim(bio.matrix))
  print(dim(t(vdge$E)))
  print(dim(covDesignNorm))
    
  snmDataObjOnly <- snm(raw.dat = as.matrix(vdge$E), 
                        bio.var = bio.matrix, 
                        adj.var = adj.matrix, 
                        rm.adj = TRUE,
                        verbose = TRUE,
                        diagnose = TRUE)
  snmData <- snmDataObjOnly$norm.dat
  
  exprs <- as.matrix(snmData)
  print(dim(exprs))
  adf <-  new("AnnotatedDataFrame", data=qcMetadata)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="ajusted counts by removeBatchEffect"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  return(expressionSet)
}

VoomSNM_ExprSet <- reduce_voom_SNM(x = Count_ExprSet)
```

```{r, fig.width=12, fig.height=6}
pca_fun <- function(expers_set=Count_ExprSet){
  
  # expers_set=Count_ExprSet
  
  pheno <- pData(expers_set)
  edata <- exprs(expers_set)
  pca <- prcomp(t(edata), scale. = TRUE, center = T)
  score <- inner_join(pca$x %>% data.frame() %>% 
                        rownames_to_column("SampleID")  %>%  
                        dplyr::select(c(1:3)) , 
                        pheno %>% rownames_to_column("SampleID"),
                      by = "SampleID")

  pl <- ggplot(score, aes(x=PC1, y=PC2))+
          geom_point(aes(color=Batch, shape=Organism), size=3.5)+
          guides(color=guide_legend(title=NULL,keywidth=.7, keyheight=.7))+
          scale_color_manual(values = colors)+
          theme_bw()+
          theme(axis.title = element_text(size=10, color="black", face="bold"),
                axis.text = element_text(size=9, color="black"),
                text = element_text(size=8, color="black", family="serif"),
                strip.text = element_text(size=9, color="black", face="bold", family="serif"), 
                panel.grid = element_blank(),
                legend.text=element_text(size=11, color = "black", family="serif"),
                legend.background = element_rect(color = "black", fill="white"))
  return(pl)  
}

p1 <- pca_fun(expers_set = Count_ExprSet)
p2 <- pca_fun(expers_set = VoomSNM_ExprSet)

require(cowplot)
prow <- plot_grid(
  p1 + theme(legend.position="none"),
  p2 + theme(legend.position="none"),
  align = 'vh',
  labels = c("Count", "VoomSNM"),
  hjust = -1,
  nrow = 1
)

legend <- get_legend(
  p1 + theme(legend.box.margin = margin(0, 0, 0, 12))
)
plot_grid(prow, legend, rel_widths = c(3, .4))

# cowplot::plot_grid(pca_fun(expers_set = Count_ExprSet), 
#                    pca_fun(expers_set = VoomSNM_ExprSet),
#                    ncol = 2,
#                    labels = c("Count", "VoomSNM"))

# require(patchwork)
# (pca_fun(expers_set = Count_ExprSet)) + (pca_fun(expers_set = VoomSNM_ExprSet))+
#   plot_layout(nrow = 1, guides = "collect")+
#   plot_annotation(tag_levels = "A")
```


### Function: correlation test
```{r}
cor_plot <- function(dataset=Count_ExprSet,
                     method="spearman"){
  
  # dataset=Count_ExprSet
  # method="spearman"
  
  # acquiring the mean value of group to regard as the overall expression value
  edata <- exprs(dataset)
  pheno <- pData(dataset)
  uniq_grp <- unique(pheno$Group)
  
  df_mean_value <- data.frame(NA)
  for (i in 1:length(uniq_grp)) {
    # print(i)
    phe <- pheno[pheno$Group %in% uniq_grp[i], ] 
    prf <- edata[, colnames(edata)%in%rownames(phe)] %>%
      as.matrix()
    
    mean_temp <- apply(prf, 1, mean) %>% 
      data.frame() %>%
      setNames(uniq_grp[i]) %>%
      rownames_to_column("GeneID")
    df_mean_value <- cbind(df_mean_value, mean_temp)
  }
  
  df_mean_value_cln <- df_mean_value %>% 
    dplyr::select(c("NC", "NICM", "ICM", "HCM", "DCM",
                         "TAC_NC", "TAC", "HF_NC", "HF"))
    
  rownames(df_mean_value_cln) <- df_mean_value$GeneID
  require(Hmisc)
  res <- rcorr(as.matrix(df_mean_value_cln), type = method)
  
  library(corrplot)
  return(
    corrplot(res$r, method = "circle", 
           col = colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "white",
                           "cyan", "#007FFF", "blue", "#00007F"))(100),
           cl.lim = c(0, 1))
  )
}
```


### **Figure 1B** : Pearson correlation of Gene expression among samples types. Larger and darker circles depict a higher correlation between samples.
```{r}
cor_plot(dataset = Count_ExprSet, method = "spearman") 
```


### version
```{r}
sessionInfo()
```


### Reference

1. [Estimate an overall correlation from the correlations within subgroups](https://stats.stackexchange.com/questions/132851/estimate-an-overall-correlation-from-the-correlations-within-subgroups)

2. [Cowplot: shared legends](https://wilkelab.org/cowplot/articles/shared_legends.html)
