---
title: "Acquiring the final gene express profile and phenotype"
date: "Created: 2021-03-16 Updated: `r Sys.Date()`"
output: 
  html_notebook:
    codes: hide
---


### Introduction 

After removing the unsuitable samples including `SRR915775` and `SRR9331234`, we finally acquired the proper profile and phenotype in this study. The data will be used for further investigation in the following data analysis.

1. rebuilding the new ExpressionSet Object

2. reanalysing the Differential expression analysis 


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(stringr)
library(data.table)
library(tibble)
library(ggplot2)

rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)
```


### load data
```{r}
hsp.set <- readRDS("../../Result/Profile/final/Homo_sapiens_ExprSet_VoomSNM_symbol.RDS")
mus.set <- readRDS("../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Hsymbol.RDS")

phen <- read.csv("../../Result/Phenotype/phenotype_final.csv") 
```


### The number of samples per group


1. Human: the number of samples (n=225)

  * NC: 69
  
  * DCM: 63
  
  * HCM: 8
  
  * ICM: 65
  
  * NICM: 20 
  

2. Mouse: the number of samples (n=85)
  
  * TAC_NC: 20 
    
  * TAC: 22 
    
  * HF_NC: 18
    
  * HF: 25


### get profile 

```{r}
get_profile <- function(expers_set=hsp.set,
                        metadata=phen){
  
  # expers_set=hsp.set
  # metadata=phen
  
  
  edata <- exprs(expers_set)
  sid <- intersect(metadata$SampleID, colnames(edata))
  
  datphe <- metadata %>% filter(SampleID%in%sid) %>%
      column_to_rownames("SampleID")
  datprf <- edata %>% data.frame() %>% 
      dplyr::select(rownames(datphe)) %>%
      as.matrix()  
  print(dim(datprf))
  exprs <- datprf
  adf <-  new("AnnotatedDataFrame", data=datphe)
  experimentData <- new("MIAME",
        name="ShuiLin Liao", lab="Dong gdl Lab",
        contact="dong_ming@grmh-gdl.cn",
        title="Heart-failure Experiment",
        abstract="The gene ExpressionSet",
        url="www.grmh-gdl.cn",
        other=list(notes="Final Profile matrix"))
  expressionSet <- new("ExpressionSet", exprs=exprs,
                       phenoData=adf, 
                       experimentData=experimentData)
  
  return(expressionSet)
}


hsp.set.new <- get_profile(expers_set=hsp.set, metadata=phen)
mus.set.new <- get_profile(expers_set=mus.set, metadata=phen)
```


### output 
```{r}
if(!dir.exists("../../Result/Profile/final/")){
  dir.create("../../Result/Profile/final/")
}

saveRDS(hsp.set.new, "../../Result/Profile/final/Homo_sapiens_ExprSet_VoomSNM_symbol_final.RDS", compress = TRUE)
saveRDS(mus.set.new, "../../Result/Profile/final/Mus_musculus_ExprSet_VoomSNM_Hsymbol_final.RDS", compress = TRUE)
```


### version
```{r}
sessionInfo()
```

